\section{Method}
\label{sec:method}

We consider the problem of learning a directed coupling network from perturbed
steady-state observations of a nonlinear dynamical system.  The method is
designed for settings in which repeated interventions can be applied, the system
can be allowed to relax, and the resulting equilibrium state can be measured
more reliably than time derivatives.  The statistical object of interest is not
the full transient trajectory, but the collection of univariate steady-state
effects that explain how perturbations propagate through the system.

\subsection{A sparse steady-state-compatible network model}
\label{sec:method:model}

Let $\xv(t)=(x_1(t),\ldots,x_p(t))^\top\in\R_+^p$ denote the state of $p$
interacting variables.  We model perturbed network dynamics through
\begin{equation}
  \frac{d x_j(t)}{dt}
  =
  q_j\{x_j(t)\}
  \left[
  \mu_j
  + f_{jj}\{x_j(t)\}
  + \sum_{i\neq j} f_{ji}\{x_i(t)\}
  + u_j(t)
  \right],
  \qquad j=1,\ldots,p,
  \label{eq:ode}
\end{equation}
as a steady-state-compatible working model.  Here $q_j(a)>0$ on the analysed
state range is a target-specific activity or turnover factor, $\mu_j$ is a
baseline drift term, $u_j(t)$ is an externally controlled perturbation, $f_{jj}$
is the self-feedback function for node $j$, and $f_{ji}$ for $i\neq j$ is the
effect of source node $i$ on target node $j$.

The form in \eqref{eq:ode} is a separable member of the broad network-dynamics
classes used to study steady-state perturbation response and resilience in
complex systems \citep{barzel2013universality,gao2016universal}.  The special
case $q_j\equiv 1$ gives an additive ODE.  The generalized Lotka--Volterra
(gLV) model corresponds to $q_j(x_j)=x_j$ with linear cross-effects and linear
self-damping, and is widely used in ecological and microbiome dynamics
\citep{stein2013ecological,venturelli2018deciphering}.  PSS-Net does not
estimate $q_j$; it estimates the additive bracket whose zero defines the
perturbed steady-state relation.

The directed network is defined by nonzero cross-node effects,
\begin{equation}
  E = \{(j\leftarrow i): f_{ji}\not\equiv 0,\ i\neq j\}.
  \label{eq:edge}
\end{equation}
Self-feedback functions are estimated because they affect equilibrium and local
stability, but they are not counted as directed network edges.

The decomposition inside the bracket in \eqref{eq:ode} has the usual additive
identifiability issue:
a constant can be moved between $\mu_j$ and any $f_{ji}$.  We impose the
normalization
\begin{equation}
  f_{ji}(0)=0,\qquad i,j=1,\ldots,p,
  \label{eq:zero}
\end{equation}
so that all constant offsets are absorbed by $\mu_j$.  This convention also
matches the interpretation that a source variable at the zero reference level
contributes no additional effect.

\subsection{Perturbed steady-state observations}
\label{sec:method:pss}

Suppose that $N$ perturbation conditions
$\uv^{(1)},\ldots,\uv^{(N)}\in\R^p$ are applied.  For each condition, the system
is allowed to reach a stable equilibrium $\xv^{(k)*}$, and we observe
\begin{equation}
  \tilde{\xv}^{(k)} = \xv^{(k)*} + \bm{\eta}^{(k)},
  \qquad k=1,\ldots,N,
  \label{eq:obs}
\end{equation}
where $\bm{\eta}^{(k)}$ represents measurement noise.  At an equilibrium with
positive activity factor $q_j\{x_j^{(k)*}\}$, \eqref{eq:ode} implies the
algebraic constraint
\begin{equation}
  \mu_j
  + f_{jj}\{x_j^{(k)*}\}
  + \sum_{i\neq j} f_{ji}\{x_i^{(k)*}\}
  + u_j^{(k)}
  = 0,
  \qquad k=1,\ldots,N.
  \label{eq:ss}
\end{equation}
Thus, for each target node $j$, the negative perturbation $-u_j^{(k)}$ acts as
the response and the measured steady-state variables provide the predictors.
This is the key statistical reduction: estimation can proceed by regression
without numerical differentiation of noisy trajectories
\citep{meister2013pss,henderson2014network,wu2014saode}.

For a multiplicative generalized Lotka--Volterra system
$\dot x_j = x_j\{r_j+\sum_{i\neq j}A_{ji}x_i-\gamma_j x_j+u_j\}$, positive
equilibrium similarly forces the bracket to vanish, which is exactly
\eqref{eq:ss} with the linear effects $f_{ji}(x_i)=A_{ji}x_i$, self-feedback
$f_{jj}(x_j)=-\gamma_j x_j$, and $\mu_j=r_j$.  PSS-Net therefore recovers the
interaction signs and supports encoded in the steady-state equation even when
the transient vector field contains an unestimated multiplicative prefactor.

\subsection{Basis expansion}
\label{sec:method:basis}

Following the dictionary-of-functions view of data-driven dynamics
\citep{brunton2016sindy,kaptanoglu2022pysindy}, we approximate each unknown
univariate effect by a finite dictionary
$\psiv(a)=(\psi_1(a),\ldots,\psi_M(a))^\top$,
\begin{equation}
  f_{ji}(a)
  \approx
  \psiv(a)^\top \thetav_{ji}
  =
  \sum_{m=1}^{M} \theta_{jim}\psi_m(a),
  \qquad
  \thetav_{ji}\in\R^M.
  \label{eq:basis}
\end{equation}
The estimator only requires a no-intercept dictionary, so that
$\psi_m(0)=0$ and the fitted effects obey the normalization in
\eqref{eq:zero}.  The default implementation uses the monomial dictionary
$\psi_m(a)=a^m$, $m=1,\ldots,M$, because it keeps linear and higher-order
contributions separated in the coefficient vector.  Other no-intercept
dictionaries, such as saturating or Fourier-type libraries, can be substituted
without changing the grouped regression structure.

For target $j$, define the stacked coefficient vector
\[
  \thetav_j
  =
  (\thetav_{j1}^\top,\ldots,\thetav_{jp}^\top)^\top
  \in \R^{pM},
\]
and construct the design matrix $\Psib\in\R^{N\times pM}$ with entries
\begin{equation}
  \Psib_{k,(i-1)M+m}
  =
  \psi_m\!\left(\tilde{x}_i^{(k)}\right),
  \qquad
  k=1,\ldots,N,\ i=1,\ldots,p,\ m=1,\ldots,M.
  \label{eq:design}
\end{equation}
Stacking \eqref{eq:ss} over the $N$ perturbation conditions gives the node-wise
linear working model
\begin{equation}
  \Psib\thetav_j + \mu_j\mathbf{1}_N
  =
  -\uv_{\cdot j} + \bm{\varepsilon}_j,
  \label{eq:linear}
\end{equation}
where $\uv_{\cdot j}=(u_j^{(1)},\ldots,u_j^{(N)})^\top$ and
$\bm{\varepsilon}_j$ collects measurement noise and basis truncation error.

\subsection{Centering, scaling, and optional intercept recovery}
\label{sec:method:preprocess}

Because the basis columns generally have nonzero empirical means, the intercept
and the effect coefficients are estimated more stably after centering.  Let
$\bar{\Psib}=N^{-1}\mathbf{1}_N^\top\Psib$ and
$\bar u_j=N^{-1}\mathbf{1}_N^\top \uv_{\cdot j}$.  Define
\begin{equation}
  \Psic = \Psib-\mathbf{1}_N\bar{\Psib},
  \qquad
  \uv_{c,j} = \uv_{\cdot j}-\bar u_j\mathbf{1}_N.
  \label{eq:center}
\end{equation}
The centered regression is
\begin{equation}
  \Psic\thetav_j
  =
  -\uv_{c,j} + \bm{\varepsilon}_j.
  \label{eq:centered-linear}
\end{equation}
For numerical conditioning and fair penalization across basis columns, each
column of $\Psic$ is divided by its empirical standard deviation.  Let
$s_\ell=\operatorname{sd}(\bm{\Psi}_{c,\cdot\ell})$ and
$s_\ell\leftarrow \max(s_\ell,10^{-10})$ for protection against constant
columns.  With $S=\operatorname{diag}(s_1,\ldots,s_{pM})$, the scaled design is
$\Psic^{(s)}=\Psic S^{-1}$.  If $\hat{\thetav}_j^{(s)}$ is estimated on the
scaled design, the original-scale estimate is
\begin{equation}
  \hat{\thetav}_j = S^{-1}\hat{\thetav}_j^{(s)}.
  \label{eq:unscale}
\end{equation}
When the full steady-state bracket rather than only the edge set or local
effects is needed, the baseline term can be recovered from the uncentered
estimating equation:
\begin{equation}
  \hat\mu_j = -\bar u_j - \bar{\Psib}\hat{\thetav}_j.
  \label{eq:intercept}
\end{equation}

\subsection{Double-sparse estimation}
\label{sec:method:adsiht}

The coefficient vector has two natural sparsity structures.  First, the network
is sparse: for a fixed target node $j$, only a small number of source nodes have
nonzero cross-node effects.  Second, each active univariate effect is expected to
have a simple representation in the chosen dictionary.  Formally, we use
\begin{align}
  \left|\{i\neq j:\norm{\thetav_{ji}}_2>0\}\right| &\le s,
  \label{eq:sparse-group}\\
  \norm{\thetav_{ji}}_0 &\le s_0
  \qquad\text{for active } (j\leftarrow i).
  \label{eq:sparse-within}
\end{align}

Our primary estimator is adaptive double sparse iterative hard thresholding
(\adsiht{}; \citealp{zhang2023minimax}), chosen to match the source-level and
within-source sparsity in \eqref{eq:sparse-group}--\eqref{eq:sparse-within}
\citep{cai2022sparsegroup,zhang2023minimax}.  For each target node $j$, it
solves the scaled version of
\begin{equation}
  \min_{\thetav_j}
  \frac{1}{2N}
  \norm{\Psic\thetav_j+\uv_{c,j}}_2^2
  \quad
  \text{subject to \eqref{eq:sparse-group}--\eqref{eq:sparse-within}}.
  \label{eq:adsiht-objective}
\end{equation}
The groups are defined by source node, so that the $M$ basis coefficients of a
given source are selected or dropped together.  The information criterion in
\adsiht{} selects the effective sparsity levels adaptively, avoiding manual
tuning of $(s,s_0)$.
When explicitly comparing node-wise and joint recovery, we also use a
multi-task version that stacks all target regressions and encourages shared
model selection across equations \citep{dong2026multitask}.

As a continuous-shrinkage comparator, we also consider the group lasso estimator
\citep{lounici2011group},
\begin{equation}
  \min_{\thetav_j}
  \left\{
  \frac{1}{2N}\norm{\Psic^{(s)}\thetav_j^{(s)}+\uv_{c,j}}_2^2
  + \lambda \sum_{i=1}^{p}\norm{\thetav_{ji}^{(s)}}_2
  \right\}.
  \label{eq:group-lasso}
\end{equation}
Unlike hard-thresholding, group lasso typically does not produce exact group
zeros, so edge selection requires a post-estimation threshold.

\subsection{Network reconstruction and local linear effects}
\label{sec:method:network}

For \adsiht{}, the native estimated support is the set of source groups selected
by the minimum-DSIC solution.  Let $\widehat{\mathcal A}_j$ be the selected
basis-coordinate set for target $j$, and let
$G_i=\{(i-1)M+1,\ldots,iM\}$ denote the source-$i$ group.  The native PSS-Net
edge set is
\begin{equation}
  \hat E
  =
  \{(j\leftarrow i): i\neq j,\ \widehat{\mathcal A}_j\cap G_i\neq\emptyset\}.
  \label{eq:edge-estimate}
\end{equation}
Group norms, fitted Jacobian magnitudes, or normalized local-response scores can
also be used to rank candidate edges.  Any post-estimation threshold used for a
continuous-shrinkage comparator or for a scale-matched external benchmark is an
analysis-specific choice and is reported with that simulation.

To summarize the local direction and magnitude of effects, we evaluate the
Jacobian of the fitted additive steady-state bracket at an analysis-specified
reference state $\xv^{\mathrm{ref}}$:
\begin{equation}
  \hat J_{ji}
  =
  \left.
  \frac{\partial}{\partial x_i}
  \left[
  \hat\mu_j + \hat f_{jj}(x_j)
  + \sum_{i'\neq j}\hat f_{ji'}(x_{i'})
  \right]
  \right|_{\xv=\xv^{\mathrm{ref}}}
  =
  \psiv'\!\left(x_i^{\mathrm{ref}}\right)^\top \hat{\thetav}_{ji}.
  \label{eq:jacobian}
\end{equation}
For the monomial dictionary, this reduces to
\begin{equation}
  \hat J_{ji}
  =
  \sum_{m=1}^{M}
  m\left(x_i^{\mathrm{ref}}\right)^{m-1}\hat\theta_{jim}.
  \label{eq:jacobian-monomial}
\end{equation}
A positive $\hat J_{ji}$ indicates that increasing node $i$ locally increases
the fitted steady-state bracket for node $j$ near the reference state, whereas a
negative value indicates a locally decreasing effect.  When $q_j\equiv 1$, this
is also the local vector-field Jacobian; with a positive multiplicative
prefactor such as gLV, the off-diagonal transient Jacobian at equilibrium is
scaled by $q_j(x_j^{\mathrm{ref}})$ and has the same sign.

\subsection{The PSS-Net algorithm}
\label{sec:method:algorithm}

Algorithm~\ref{alg:pssnet} collects the preceding steps.  Because the design
matrix $\Psib$ is shared across all target nodes---only the response
$\uv_{\cdot j}$ changes with $j$---the $p$ regressions can be written as a single
global double-sparse problem with the Kronecker design $\Xb=\Ib_p\otimes\Psic^{(s)}$
and stacked response $\yv=-\operatorname{vec}(\Uc)$, where
$\Uc\in\R^{N\times p}$ collects the centered perturbations.  The block-diagonal
grouping then has one group per $(\text{target},\text{source})$ pair.  Because
the least-squares loss is separable across targets, the node-wise loop is the
memory-efficient default and is exactly equivalent when each target is selected
independently.  The block-diagonal form differs only when a single selection
criterion is applied across all blocks: pooling the model-complexity decisions
couples information across tasks and empirically improves recovery on
heterogeneous, scale-free networks, where a few hub targets carry high in-degree
while most carry little.  This benefit comes at substantially higher memory
cost---the dense Kronecker design scales with $p^2$---so the joint form is
preferred for hub-dominated structure with small $N$ and the node-wise loop
elsewhere.

\begin{algorithm}[t]
\caption{PSS-Net: directed network inference from perturbed steady states}
\label{alg:pssnet}
\begin{algorithmic}[1]
\Require Steady-state observations $\{\tilde{\xv}^{(k)}\}_{k=1}^N$, perturbations
  $\{\uv^{(k)}\}_{k=1}^N$, no-intercept dictionary $\psiv$ of size $M$,
  reference state $\xv^{\mathrm{ref}}$
\Ensure Directed edge set $\hat E$, local bracket Jacobian $\hat{\Jac}$
\State Build design matrix $\Psib$ with
  $\Psib_{k,(i-1)M+m}=\psi_m(\tilde x_i^{(k)})$ \Comment{Eq.~\eqref{eq:design}}
\State Center and column-scale:
  $\Psic=\Psib-\mathbf 1_N\bar\Psib$, $\;\Psic^{(s)}=\Psic S^{-1}$
  \Comment{Eqs.~\eqref{eq:center}--\eqref{eq:unscale}}
\State Form $\Xb=\Ib_p\otimes\Psic^{(s)}$ and $\yv=-\operatorname{vec}(\Uc)$
\State $\hat\thetav^{(s)}\gets\adsiht{}\big(\Xb,\yv;\,
  \texttt{group}=\mathrm{rep}(1{:}p^2,\ \text{each}{=}M)\big)$
  \Comment{double-sparse, IC-tuned \eqref{eq:adsiht-objective}}
\State Unscale and reshape to $\{\hat\thetav_{ji}\}_{i,j=1}^p$; recover
  $\hat\mu_j$ if the full bracket is needed \Comment{Eq.~\eqref{eq:intercept}}
\State Edge set $\hat E$ from the selected source groups \Comment{Eq.~\eqref{eq:edge-estimate}}
\State Bracket Jacobian
  $\hat J_{ji}=\psiv'(x_i^{\mathrm{ref}})^\top\hat\thetav_{ji}$
  \Comment{Eq.~\eqref{eq:jacobian}}
\State \Return $\hat E,\ \hat{\Jac}$
\end{algorithmic}
\end{algorithm}

\subsection{Application datasets}
\label{sec:method:data}

We apply PSS-Net to synthetic human gut microbiome communities, which match the
perturbed steady-state design: defined species combinations are assembled,
cultured to a community endpoint, and the resulting abundances are read out.
The primary dataset is the 26-species assembly of \citet{clark2021design}
($N=1850$ species combinations measured by 16S amplicon sequencing), with the
12-species, OD600-calibrated communities of \citet{venturelli2018deciphering} as
a smaller, near-absolute-abundance secondary set.  Species presence/absence (or
inoculation level) plays the role of the perturbation $\uv$ and the endpoint
abundances play the role of the steady state $\tilde{\xv}$.  Because 16S
measurements are compositional, abundances are normalized by a centered
log-ratio or reference-species transform before inference, except where an
external total-abundance measurement (OD600) permits an absolute-scale analysis;
the impact of this choice is quantified in the compositional simulation
(Section~\ref{sec:sim:benchmark}).
